The Evolution of Phenotypically Invariant Genetic Networks

### 2017-10-20 kalman_walk_paper.tex
more edits from overleaf

\newcommand{\calA}{\mathcal{A}}
\newcommand{\calH}{\mathcal{H}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

\newcommand{\Sys}{\mathcal{S}}
\newcommand{\allS}{\mathcal{A}}
%\newcommand{\allS}{\mathcal{A}}

@@

\plr{Additional ideas to consider adding:}
\begin{itemize}
    \item speciation literature
    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    %\item hybrid vigor (need to do calculation)
    \item discuss linearity, linearization, and canalization in introduction
    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
\end{itemize}

@@

{Linearity} also ``zero-state equivalence''
Here we have assumed that the system is linear,
and begins from the ``zero'' state ($\kappa(0)=0$).
Of course, neither of these are necessarily true for real systems,
but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
Furthermore, the ease of analyzing linear systems makes this an attractive place to start.

@@

      Suppose the input is an impulse at time zero (a delta function), and so its phenotype is equal to its impulse response,
      \begin{align*}
        \phi(t) = h(t) = \sin t + \cos t  .
      \end{align*}
      The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
    \begin{figure}[H]
       % \begin{center} 
      \centering
         \begin{tabular}{cc}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$\kappa_{1}$};
                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
                \node[shape=rectangle] (U) at (1.5,1.5) {input};
                \node[shape=rectangle] (y) at (1.5,-1.5) {output};
            \end{scope}

@@

%Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype. Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes. These changes cannot be described by $\mathcal{S}(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.
%The Kalman decomposition decomposes a system into \emph{reachable} and \emph{observable} components -- that is coefficients that respond to an input and coefficients that produce an observable output. A minimal system's components are both reachable and observable. In this form, it is possible to add components that respond to an input, but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs. Any system with such components is, by definition, non-minimal, and any system can be modified to include any finite number of extra unreachable and/or unobservable components, without modifying its impulse response. Decomposing the system allows for the easy manipulation of ``superfluous''(but not necessarily expendable) components. After unreachable/unobservable coefficients are added, changing coordinates ``stirs'' the system coefficients such that the previously superfluous coefficients can become essential to system funciton. 
%Whether or not evolution constrains biological systems to be minimal or nearly minimal is an open question.

Two systems are equivalent if they produce the same phenotype given the same input,
i.e., have the same input--output relationship.
For any input $u(t)$,
suppose that $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ are the solutions to equation \eqref{eqn:system}
with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively.
% (recall that $\kappa(0) = \bar \kappa(0) = 0$). 
We say that
the systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
\textbf{phenotypically equivalent} 
if
\begin{align*}
      \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
\end{align*}
This occurs if and only if their impulse response functions are the same:
if $h(t) = \bar h(t)$ for all $t \ge 0$.

One way to find other systems phenotypically equivalent to a given one
is by change of coordinates:
if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
are phenotypically equivalent because their impulse response functions are equal:
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B} .
    \end{aligned}
  \end{equation}
However, not all phenotypically equivalent systems are of this form:
systems can have identical impulse responses without being coordinate changes of each other.
In fact, systems with identical impulse responses can involve interactions between different
numbers of molecules, and thus 
have kryptotypes in
different dimensions altogether.

@@

If the system is not minimal, then the Kalman decomposition tells us XXX

We refer to $\calA(\Sys_0)$ as the set of all systems equivalent to $\Sys_0$, regardless of dimension:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA(\Sys_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  .
  \end{aligned}
\end{equation}

The set of all systems phenotypically equivalent to a given system $(A,B,C)$ 
is elegantly described using the Kalman decomposition,
which also clarifies the system dynamics.
To motivate this, first note that the input $u(t)$ only directly pushes the system
in directions lying in the span of the columns of $B$.
As a result, different combinations of input can 
move the system in any direction that lies in the \emph{reachable subspace},
which we denote by $\reachable$,
and is defined to be the closure of $\spn(B)$ under applying $A$
(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
Analogously to this, we define
the \emph{observable subspace}, $\mathcal{O}$,
to be the closure of $\spn(C^T)$ under applying $A$.
(Or: $\unobservable$ is the largest $A$-invariant subspace
contained in the null space of $C$;
and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)

\begin{theorem}[Kalman decomposition] \label{thm:kalman}
        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
        \begin{align*}
            \widehat A = PAP^{-1}
            &=
            \left[ \begin{array}{cccc}
                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
                0 & 0 & 0 & A_{\nro}
            \end{array} \right] ,
        \end{align*}
        and
        \begin{align*}
            \widehat B = PB
            &=
            \left[ \begin{array}{cccc}
                B_{\rno} \\
                B_{\ro} \\
                0 \\
                0 
            \end{array} \right] ,
        \end{align*}
        and
        \begin{align*}
            \widehat C = CP^{-1}
            &=
            \left[ \begin{array}{cccc}
                0 & C_{\ro} & C_{\nrno} & 0 
            \end{array} \right] .
        \end{align*}
        The impulse response of both systems is given by
        \begin{align*}
          h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro} .
        \end{align*}
\end{theorem}

In the latter case, we say that the system is \emph{minimal} 
-- there is no equivalent system with a smaller number of molecular species.
Note that this says that any two equivalent minimal systems
are changes of coordinates of each other.

Since any system can be put into this form,
and once in this form, its impulse response is determined only by 
$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
therefore, the set of all equivalent systems are parameterized by
the dimension $n$,
the choice of basis ($P$),
the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
(which are unconstrained),
and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.

%Usually, the dimension $n$ and the reference system $\Sys_0$ is implicit and we write only $\calA$.
%Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
%  \begin{align}
%    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
%  \end{align}
It is remarkable to note that even with the relationships between environment, kryptotype, and phenotype constant, and in the minimal dimension, there are still almost always degrees of freedom. These correspond to distinct genetic networks that perform indistinguishable functions. For example, the equivalent systems in Example \ref{ex:all_osc} are minimal, and share common B and C matrices.

It is also remarkable to note that there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, there is always more than one possible gene regulatory network architecture per phenotype. 
%This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} (Appendix \ref{apx:kalman}).

For the remainder of the paper, we interpret $\allS$ as the phenotypially neutral landscape, wherein a large population will drift under environmental and selective stasis. Even if the phenotype is constrained and remains constant through evolutionary time, the molecular mechanism underpinning it is not constrained and likely will not be conserved.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection*{Speciation}

%Define reproduction in diploids as first, the recombination of unlinked genes to make gametes, and second, as the averaging of two individual parental gametes to produce an offspring. Assuming parental populations are both phenotypically identical and genetically homogenous within each population, first generation hybrids (F1s) can be computed by averaging the two parental gene networks. Second generation hybrids (F2s) can be computed by first swapping the genes between the two parental gene networks, and next averaging these hybrid gametes.

%Speciation, or the formation of new species, is often a consequence of reprdouctive incompatibility among populations -- usually via hybrid sterility or inviability. To discuss speciaiton we first need to describe how reproduction is modelled.
 %A diploid first generation hybird's ($F_1$'s) genetic regulation is a consequence of its two genomes, as a copy of genes and regulatory sequences from each parent is present.
%\plr{should have a stronger justification for the assertion that $F_1$ is the arithmetic average of the parents than "Thus we say"}
%A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks. 
%\plr{I don't think we say that $A$ is a "gene network architecture".}
%Therefore the phenotypic dynamics typical of an $F_{1}$ cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid crosses $F_2$ between allopatric populations, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.

%\begin{align}
%    F_{1}(\tau, \hat{\tau}) &= \frac{ A(\tau) + A(\hat{\tau}) }{2}  .
%   h(t, F_{1}) &= C e^{\left( F_{1}\left( \tau, \hat{\tau} \right) t \right)} B
%\end{align}
%\begin{align*}
% G\left(r, \tau, \hat{\tau}\right) = Q(r) A( \tau ) + \left( I - Q(r) \right) A( \hat{\tau} )
%\end{align*}
%\begin{align}
%    F_{2}(i, j) = \frac{G (i) + G(j)}{2}
%\end{align}

  %\begin{align*}
  %  h\left(t, F_{2}\left( r, r' \right)\right) = C e^{\left( F_{2}\left(r, r' \right) t \right)} B
  %\end{align*}

%Fitness $\fit$ is a function (usually a Guassian) of an organism's weighted phenotypic distance from an optimum,
%  $\fit\left( \widehat{\phi}(t) \right) = \exp \left(- \int_{0}^{t} \frac{\rho(s)}{\beta} \left\lVert \phi(s) - \widehat{\phi}(s) \right\rVert^{2} ds  \right)$
%where $\rho(t)$ is some weighting function and $\beta$ is a parameter.

%We quantitatively describe the degree of incompatibility between two populations $P_{1}$ and $P_{2}$ as
%\begin{align*}
%      \Incompat = \frac{2 \left\langle \fit \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \fit \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \fit \left(\phi_{P_{2}}\right) \right\rangle} ,
%\end{align*}
%where angled brackets imply averaging.  \plr{averaging over what? And, this is as others define - cite.}

%\plr{This is not part of the definition of $\mathcal{I}$:}
%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
%%\plr{don't need these to be big equations.}
%%\begin{align*}
%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
%%  \mathcal{I} &= 0
%%\end{align*}
%
%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
%
%  \begin{figure}[H]
%    \centering
%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
%  \end{figure}
%  }
%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
%%       DMI examples...
%%       \begin{figure}[H]
%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
%%       \end{figure}
%%     \end{example}
%
%%     \begin{example}[Not all Networks can Host Incompatibilities]
%%       convex sets cant have DMIs
%
%%       \begin{align*}
%%         h(t) = 2 e^{- \theta t}
%%       \end{align*}
%%       Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
%%     \end{example}
%
%
%\plri{Missing: a statement about when there are neutral directions, and how many.}
%
%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
%
%
 \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
   The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by $\Sys (\tau) = (A(\tau), B, C)$, such that, 
    %\begin{align*}
    %  A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
    %\end{align*}
    \begin{align*}
      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
% Nice tikz!!!
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
\end{scope}

\end{tikzpicture}
      \caption{$\Sys(\tau)$, the set of all phenotype-equivalent cell cycle control networks $\Sys (\tau) = \left\{ \calA_2 : B = [1\ 1]^{T} ,\ C = [1\ 0] \right\}$.}
    \end{figure}
 % \end{example}

 % \begin{example}[External Equivalence does not imply internal equivalence]
   Despite the phenotypic equivalence of all instantiations of $\Sys(\tau)$, the kryptotypes, vary as a function of $\tau$. 
    %Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
  \begin{figure}[H]
    \centering
   % \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}

>>>>

\newcommand{\calH}{\mathcal{H}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

% the triple (A,B,C) as 'system' (not using)
\newcommand{\Sys}{\mathcal{S}}
% neutral set of all systems
\newcommand{\allS}{\mathcal{N}}

@@

\plr{Additional ideas to consider adding:}
\begin{itemize}
    \item speciation literature
    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    %\item hybrid vigor (need to do calculation)
    \item discuss linearity, linearization, and canalization in introduction
    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
    \item ``On the origin of species not by means of natural selection''
\end{itemize}

@@

% {Linearity} also ``zero-state equivalence''
Here we have assumed that the system is linear,
and begins from the ``zero'' state ($\kappa(0)=0$).
Of course, neither of these are necessarily true for real systems,
but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
Furthermore, the ease of analyzing linear systems makes this an attractive place to start.

@@

      Suppose the input is an impulse at time zero (a delta function), and so its phenotype is equal to its impulse response,
      \begin{align*}
        \phi(t) = h(t) = \sin t + \cos t  .
      \end{align*}
      The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
    \begin{figure}[H]
       % \begin{center} 
      \centering
         \begin{tabular}{cc}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$\kappa_{1}$};
                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
                \node[shape=rectangle] (U) at (1.5,1.5) {input ($u(t)$)};
                \node[shape=rectangle] (y) at (1.5,-1.5) {output ($\phi(t)$)};
            \end{scope}

@@

Two systems are equivalent if they produce the same phenotype given the same input,
i.e., have the same input--output relationship.
We say that
the systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
\textbf{phenotypically equivalent} 
if their impulse response functions are the same:
$h(t) = \bar h(t)$ for all $t \ge 0$.
This implies that for any acceptable input $u(t)$,
if $(\kappa_u(t),\phi_u(t))$ and $(\bar \kappa_u(t),\bar \phi_u(t))$ are the solutions to equation \eqref{eqn:system}
of these two systems, respectively, then
\begin{align*}
      \phi_u(t) = \bar \phi_u(t) \qquad \text{for all} \; t \ge 0.
\end{align*}
% (recall that $\kappa(0) = \bar \kappa(0) = 0$).

One way to find other systems phenotypically equivalent to a given one
is by change of coordinates:
if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
are phenotypically equivalent because their impulse response functions are equal:
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B} = \bar h(t).
    \end{aligned}
  \end{equation}
However, not all phenotypically equivalent systems are of this form:
systems can have identical impulse responses without being coordinate changes of each other.
In fact, systems with identical impulse responses can involve interactions between different
numbers of molecules, and thus 
have kryptotypes in
different dimensions altogether.

@@

Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, 
and also still preserve their phenotypes.
More generally, even if the system is not minimal, 
results from systems theory % -- the Kalman decomposition --
explicitly describe the set of all phenotypically equivalent systems.
% Could define script S here for 'system' if we want: $\Sys$
We refer to $\allS(A_0,B_0,C_0)$ as the set of all systems phenotypically equivalent
to the system defined by $(A_0, B_0, C_0)$.
Concretely, this is
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \allS(A_0, B_0, C_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  .
  \end{aligned}
\end{equation}
These systems need not have the same kryptotypic dimension $n$,
but must have the same input and output dimensions ($\ell$ and $m$, respectively).

The Kalman decomposition, which we now describe informally, elegantly characterizes this set
\citep{kalman1963mathematical,kalman1969topics,anderson1966equivalence}.
To motivate this, first note that the input $u(t)$ only directly pushes the system
in certain directions (those lying in the span of the columns of $B$).
As a result, different combinations of input can 
move the system in any direction that lies in what is known as the \emph{reachable subspace}.
Analogously, we can only observe motion of the system in certain directions
(those lying in the span of the columns of $C$),
and so can only infer motion in what is known as the \emph{observable subspace}.
The Kalman decomposition then classifies each direction in kryptotype space
as either reachable or unreachable, and as either observable or unobservable.
Only the components that are both reachable and observable determine the system's phenotype --
that is, components that respond to an input and components that produce an observable output. 
% it is possible to add components that respond to an input, 
% but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs.

Concretely, the \textbf{Kalman decomposition} of a system $(A,B,C)$  
gives a change of basis $P$ such that
the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
\begin{align*}
       PAP^{-1}
       &=
       \left[ \begin{array}{cccc}
           A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
           0 & A_{\ro} & 0 & A_{\ro,\nro} \\
           0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
          0 & 0 & 0 & A_{\nro}
       \end{array} \right] ,
\end{align*}
and
\begin{align*}
     PB
     &=
    \left[ \begin{array}{cccc}
         B_{\rno} \\
         B_{\ro} \\
         0 \\
         0 
   \end{array} \right] 
   &
   (CP^{-1})^T
   &=
   \left[ \begin{array}{cccc}
       0 \\
       C_{\ro}^T \\
       C_{\nrno}^T \\
       0 
   \end{array} \right] .
\end{align*}
% and
% \begin{align*}
%    CP^{-1}
%    &=
%    \left[ \begin{array}{cccc}
%        0 & C_{\ro} & C_{\nrno} & 0 
%    \end{array} \right] .
%\end{align*}
The impulse response of the system is given by
\begin{align*}
      h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro},
\end{align*}
and therefore, the system is phenotypically equivalent to the \emph{minimal} system $(A_{\ro}, B_{\ro}, C_{\ro})$.
(Here the subscript $\ro$ refers to the both \emph{reachable and observable} subspace,
while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
and similarly for $\nro$ and $\rno$.)

Any two minimal systems are related by a change of coordinates,
and so the minimal subsystems obtained by the Kalman decomposition
are unique up to a change of coordinates.
In particular, this implies that there is no equivalent system with a smaller number of kryptotypic dimensions
than the dimension of the minimal system.
It is also remarkable to note that the gene regulatory network architecture to achieve a given input--output map is never unique --
both the change of basis used to obtain the decomposition
and, once in this form, all submatrices other than $A_{\ro}$, $B_{\ro}$, and $C_{\ro}$ can be changed without affecting the phenotype,
and so represent degrees of freedom.

\emph{Note on implementation:}
The \emph{reachable subspace},
which we denote by $\reachable$,
is defined to be the closure of $\spn(B)$ under applying $A$,
% (or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
% Analogously to this, we define
and the \emph{unobservable subspace}, 
%$\mathcal{O}$, is the closure of $\spn(C^T)$ under applying $A$.
denoted $\unobservable$, is the largest $A$-invariant subspace
contained in the null space of $C$.
% and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
The four subspaces, $\rno$, $\ro$, $\nrno$, and $nro$
are defined from these by intersections and orthogonal complements.

%Usually, the dimension $n$ and the reference system $\Sys_0$ is implicit and we write only $\calA$.
%Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
%  \begin{align}
%    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
%  \end{align}

For the remainder of the paper, we interpret $\allS$ as the phenotypically neutral landscape, 
wherein a large population will drift under environmental and selective stasis. 
Even if the phenotype is constrained and remains constant through evolutionary time, 
the molecular mechanism underpinning it is not constrained and likely will not be conserved.

Finally, note that if $B$ and $C$ are held constant --
i.e., if the relationships between environment, kryptotype, and phenotype do not change --
there are \emph{still} usually degrees of freedom. 
These correspond to distinct genetic networks that perform indistinguishable functions. 
The following example \ref{ex:all_osc} gives the set of minimal systems equivalent to the oscillator of example \ref{ex:oscillator},
that all share common $B$ and $C$ matrices.

\begin{example}[All Phenotypically Equivalent Oscillators] \label{ex:all_osc}
The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
by an invertible matrix $V$.
If we further require $B$ and $C$ to be invariant then we need $VB=B$ and $CV=C$.
Solving these equations,
we find that a one-parameter family $(A(\tau), B, C)$ describes the set of all two-gene systems
phenotypically equivalent to the oscillator,
where
    \begin{align*}
      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1 .
    \end{align*}
The resulting set of systems,
and their dynamics, are depicted in Figure \ref{fig:all_osc}
\end{example}

    \begin{figure}[H]
    \centering
\begin{tabular}{cc}
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
\end{scope}
\end{tikzpicture} &	\includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
    \end{tabular}
% \caption{}
      \caption{
      (Left) $A(\tau)$, the set of all phenotype-equivalent cell cycle control networks.
      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green).
      % Both gene-1 dynamics are given by $\kappa_{1}(t) = \sin t + \cos t$, and gene-2 by $\kappa_{2}(t) = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2}(t) = \cos t + 3 \sin t$ ($A(2)$).
      } 
    \label{fig:all_osc}
\end{figure}

\paragraph{Sexual reproduction and recombination}
Parents with phenotypically equivalent, yet different, gene networks may produce offspring with dramatically different phenotypes. 
If the phenotypes are significantly divergent then the offspring may be inviable or otherwise have very low fitnesses, 
despite both parents being well adapted. If this is consistent for the entire population, we would consider them to be separate species.

Diploid organisms have two copies of the genome, each of which encodes a set of system coefficients.
We assume that a diploid who has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from her mother and father, respectively,
has phenotype determined by the system that averages these two,
$((A'+A'')/2, (B'+B'')/2, (C'+C'')/2)$.

Each of the copies of the genome an organism inherits from her parents are generated from their two copies of the genome by meiosis,
in which the diploid parent recombines her two genomes.
We will assume that each coefficient (i.e., entry of $A$, $B$ or $C$)
is determined by a single nonrecombining locus,
so that each coefficient in the system produced by meiosis is an independent random choice
of the two parental coefficients.
With these definitions, an $F_1$ offspring of two individuals
carries one system copy from each parent,
and an $F_2$ is the offspring of two independently formed $F_1$ individuals.
If the parents are from distinct populations,
these are first- and second-generation hybrids, respectively.

This is a simplification:
since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
and hence is determined by the promoter region of gene $i$,
we would actually expect the elements of a row of $A$ to tend to be inherited together.
Similarly, we expect in practice heritable variation in each coefficient
to be determined by more than one locus --
but this may be a reasonable approximation.

The offspring of two systems with the same phenotype may not have the same phenotype as the parents -- in other words $\allS$, the set of all systems phenotypically equivalent to a given one, is not, in general, closed under averaging or recombination.
Next we discuss how this fact can contribute to hybrid incompatibility and genetic load.
If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.

\begin{figure}[H]
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    A conceptual figure of the fitness consequences of hybridization:
    axes represent system coefficients (i.e., entries of $A$);
    the line of optimal system coefficients is down in black;
    dotted lines give phenotypic distances to the optimum.
    Two pairs of parental populations are shown in black, along the optimum;
    a hypothetical population of $F_1$s are shown for each in red,
    and the distribution of one type of $F_2$ is shown in purple
    (other types of $F_2$ are not shown).
    Solid lines depict the distance of the $F_2$ to optimum.
}
\end{figure}

\paragraph{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
This leads to the question: 
How quickly does the hybrid phenotype break down with increasing distance between the parents?
To quantify this, we will measure how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
suppose that $h_0(t)$ is the \emph{optimal} impulse response function 
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ is explicitly worked out in Appendix \ref{apx:H_calc},
and shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry --
in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set)
with respect to the coordinates.
